---
title: "Survival Analysis for molecular subtypes of HGG"
output: 
  html_notebook:
    toc: TRUE
    toc_float: TRUE
author: Jo Lynne Rokita
date: 2022
params:
  plot_ci: TRUE
---

**Purpose:** 

Runs survival analysis models for subtypes of HGG tumors. 

## Usage 

Uses a wrapper function (`survival_analysis`) from utils folder. 

## Setup

#### Packages and functions

Read in set up script.

```{r Set up library}
library(survival)
library(ggpubr)
library(tidyverse)

# This script contains a wrapper function that can conduct the survival analyses
source(file.path("util", "survival_models.R"))

# Magrittr pipe
`%>%` <- dplyr::`%>%`
```

#### Set up files and directories

Set up output directories. 

```{r Set up directories}
input_dir <- file.path("..", "02-add-histologies", "output")
results_dir <- "output"
plots_dir <- "plots"
```

Make output directories.

```{r Make output directories}
if (!dir.exists(results_dir)) {
  dir.create(results_dir)
}
if (!dir.exists(plots_dir)) {
  dir.create(plots_dir)
}
```

Declare input file paths and useful variables

```{r Set input file paths}
metadata_file <- file.path(input_dir, "stundon_hgat_updated_hist_alt.tsv")

```

Declare output file paths for HGG/DMG all subtypes. 

```{r Set output file paths}
kap_meier_plot_file <- file.path(plots_dir, "KM_hgg_alt_subtypes.pdf")
kap_meier_model_file <- file.path(results_dir, "logrank_hgg_alt_subtypes.RDS")
```

## Import the metadata

```{r Read in metadata}
# choose HGG samples
# remove cell-lines from this analysis as well
metadata <- readr::read_tsv(
  metadata_file,
  col_types = readr::cols(
    molecular_subtype = readr::col_character())) %>%
  dplyr::filter(
      short_histology == "HGAT")  %>%
  mutate(
    OS_years = OS_days / 365.25,
    ALT_status = case_when(`alt final` == "pos" ~ "ALT +",
                           `alt final` == "neg" ~ "ALT -",
                           `alt final` == "POS"~ "ALT +",
                           `alt final` == "NEG"~ "ALT -")
  ) %>%
  # keep only relevant columns
  select(Kids_First_Participant_ID, Kids_First_Biospecimen_ID_DNA, Kids_First_Biospecimen_ID_RNA, tumor_descriptor, ALT_status, molecular_subtype, OS_days, OS_years, OS_status)
```

## Create new variable
```{r add alt status to subtype}
# first update H3 subtype
meta_updated <- metadata %>%
  dplyr::mutate(mol_sub_group = case_when(
    molecular_subtype %in% c("DMG, H3 K28", 
                             "DMG, H3 K28, TP53 loss", 
                             "DMG, H3 K28, TP53 activated") ~ "H3 K28",
    molecular_subtype %in% c("HGG, H3 G35",
                             "HGG, H3 G35, TP53 loss") ~ "H3 G35",
    molecular_subtype %in% c("HGG, H3 wildtype",
                             "HGG, H3 wildtype, TP53 loss",
                             "HGG, H3 wildtype, TP53 activated") ~ "H3 WT",
    molecular_subtype %in% c("HGG, IDH, TP53 loss",
                             "HGG, IDH, TP53 activated") ~ "IDH",
    TRUE ~ NA_character_
  ))

as.data.frame(table(meta_updated$mol_sub_group))

# add in ALT status
meta_updated$alt_subtype <- paste(meta_updated$mol_sub_group, meta_updated$ALT_status, sep = ", ")

as.data.frame(table(meta_updated$alt_subtype))
```

## Kaplan-Meier for all HGG/DMG subtypes

The [Kaplan-Meier](https://www.ncbi.nlm.nih.gov/pmc/articles/PMC3059453/) 
survival curve shows the probability of survival over time.

```{r Kaplan Meier, warning=TRUE}

kap_fit <- survival_analysis(
  metadata  = meta_updated,
  ind_var = "alt_subtype",
  test = "kap.meier",
  metadata_sample_col = "Kids_First_Biospecimen_ID_DNA"
  )
# median and 0.95CI survival days
kap_fit$original_data %>%
  write_tsv(file.path(results_dir, "kap_fit_hgg_alt.tsv"))
```
Note that upper 95% bounds (0.95UCL) which are denoted "NA" are the consequence of underlying skew in the data. The upper confidence limit is therefore not defined (or may be interpreted as infinity).

`survminer` package comes with a convenient plotting function which uses `ggplot2`
arguments. 
It requires the original Kaplain-Meier `survfit` object, so we will extract that with `$model`.

```{r Make survival plot}
surv_plot <- survminer::ggsurvplot(fit = kap_fit$model,
  data = kap_fit$original_data,
  pval = TRUE,
  risk.table = TRUE,
  xlim = c(0, 4000),
  break.time.by = 500,
  ggtheme = theme_minimal(),
  risk.table.y.text.col = TRUE,
  risk.table.y.text = FALSE,
)

surv_plot$plot <- surv_plot$plot +
  ggtitle(paste0("Kaplan Meier HGG tumors by ALT subtype")) +
  theme(legend.position = "right")

# Make this plot a combined plot
surv_plot_all_subtype <-
  cowplot::plot_grid(surv_plot[[1]], surv_plot[[2]], nrow = 2,
                     rel_heights = c(3, 2))

# Print it out here
surv_plot_all_subtype
```

Save the plot to a file. 

```{r Save survival plot}
# We can save the plot like a normal ggplot
cowplot::save_plot(filename = kap_meier_plot_file, plot = surv_plot_all_subtype, base_height = 5, base_width = 8)
# Save the model itself as well as all other output from `survival_analysis()`
readr::write_rds(kap_fit, kap_meier_model_file)
```


## Analysis: cox regression for HGG ALT subtypes

```{r cox hgg subtype}
# Run model
fit_save_model(meta_updated, 
               "alt_subtype",
               file.path(results_dir, "cox_hgg_alt_subtype.RDS"))
```

## Multivariate analysis using molecular subtype and ALT subtypes
### Interaction model - not significant
`mol_sub_group*ALT_status` 

```{r int alt molsub}
# we save this for the purpose of printing out a p-value below
int_model <- fit_save_model(meta_updated,
             "mol_sub_group*ALT_status",
             file.path(results_dir, "cox_interaction_terms_alt_molsub.RDS"),
             "multivariate"
            )
int_model
```

### Additive model: `mol_sub_group*ALT_status` 
Also not significant

```{r additive alt molsub}
# we save this for the purpose of printing out a p-value below
add_model <- fit_save_model(meta_updated,     # use the data unknown resection data for this model
             "mol_sub_group+ALT_status",
             file.path(results_dir, "cox_additive_terms_mol_sub_group_ALT_status.RDS"),
             "multivariate"
            )
add_model
```


Pairwise comparisons 
```{r cox pairwise}
# Double check the function isn't missing them - get the same result below
metadata_recode <- meta_updated %>%
      mutate(OS_status = ifelse(OS_status == "LIVING", 0, 1))
pairwise_survdiff(formula = Surv(OS_years, OS_status) ~ alt_subtype, 
                  data = metadata_recode)
```




